---
title: 'Correlation of Location, Year, Economic Status, and Demographic Information with Suicide Rate'
author: "STAT 420, Summer 2020, Kun Hu, Madeline Old, Yuwen Xiang, Patrick Willhalm"
date: '07/18/2020'
output:
  pdf_document: 
    latex_engine: xelatex
  html_document: 
    theme: readable
    toc: yes  
urlcolor: cyan
---


***
# Introduction

This file compares the number of suicides per 100,000 across categories stratified by age group, sex, country, and year. There are 27820 observations across 12 variables. Variables include:

* country
* year
* sex
* age group
* number of suicides  
* population
* suicides per 100k people
* Human Development Index (HDI)
* Gross Domestic Product (GDP)
* GDP per capita


This dataset is a compilation of data from the UN Development Program, World Bank, and the World Health Organization. We found the data online at [https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016](https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016)

Suicide is unique as a public health issue in that it is 100% preventable. Stopping deaths from cancer requires developing new cancer drugs, but suicide is a social and economic rather than technical problem. Here we intend to find the correlation among different social status (includes year, sex and age) and the economic situation with the suicide rate with an eye toward suicide prevention. Depending on social status, we may find some causal links with suicide risk.

# Methods

## Load packages and the data
```{r setup, echo <- FALSE, message <- FALSE, warning <- FALSE}
library(tidyverse)
library(knitr)
library(faraway)
library(ggpubr)
library(lmtest)

options(scipen = 1, digits = 4, width = 80)
opts_chunk$set(cache = TRUE, autodep = TRUE)

suicide <- read.csv('master.csv') %>%
  as_tibble(.)
```

## Define functions for statistical tests

Here we prepared some functions for statistical tests we are going to do below. Meanwhile, we have functions to extract R-squared and adjusted R-squared values, T-test statistics and F-test statistics from ANOVA test. Finally, we also have a fundtion to wrap all the results up into one data frame, so later we will be able to have all the results stored in the same place. The qq-plot function would generate fitted vs. residuals and Q-Q plot. 
```{r, warning=FALSE}
bp_test_decision = function(model, alpha){
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

cooks_distance_inf = function(model){
  mods_cd = cooks.distance(model)
  sum(ifelse(mods_cd > 4/length(resid(model)), 1, 0))
}

calc_loocv_rmse = function(model){
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

calc_rmse = function(actual, predicted) {
  sqrt(sum((actual - predicted)^2) / length(actual)) 
}

get_t_pval = function(model){
  summary(model)$coefficients[,"Pr(>|t|)"]
}

get_f_pval = function(model1, model2){
  anova(model1, model2)[2,"Pr(>F)"]
}

get_adj_r2 = function(model){
  summary(model)$adj.r.squared
}

get_r2 = function(model){
  summary(model)$r.squared
}

get_all <- function(model) {
  bp <- bp_test_decision(model, 0.05)
  cooks <- cooks_distance_inf(model)
  loocv <- calc_loocv_rmse(model)
  r2 <- get_r2(model)
  r2_adj <- get_adj_r2(model)
  
  results <- data.frame("BP test decision" = bp,
                        "Cook's Distance" = cooks,
                        "LOOCV RMSE" = loocv,
                        "R-squared" = r2,
                        "Adjusted R-squared" = r2_adj)
  
  row.names(results) <- deparse(substitute(model))
  
  return(results)
}

'%!in%' <- function(x,y)!('%in%'(x,y))

qq_plot <- function(model, pcol = 'grey', lcol = 'dodgerblue') {
  p1 <- ggplot() +
    geom_point(aes(x = fitted(model), y = resid(model)), color = pcol) +
    geom_hline(yintercept = 0, color = lcol) +
    labs(x = "Fitted", y = "Residuals", title = "Fitted versus Residuals") +
    theme_light()
  
  p2 <- ggplot(data.frame(y = resid(model)), aes(sample = y)) + 
    stat_qq(color = pcol) + 
    stat_qq_line(color = lcol) +
    labs(x = "Theoretical", y = "Sample", title = "Normal Q-Q Plot") +
    theme_light()
  
  ggarrange(p1, p2, ncol = 2)
}
```

## Data cleaning

In the data cleaning part, we did the following:

1. Changed the variable type of age and sex from character to numeric
2. Renamed several variables
3. For each year, we calculated the sums with different conditions
4. GDP was originally in numbers with commas inside, we removed these commas and turned the GDP to numeric
5. Factorized sex and generation

```{r}
n <- nrow(suicide)
buffer <- rep(0, n)

sex <- buffer
sex[suicide$sex == 'male'] <- 1

age <- buffer
age[suicide$age == '5-14 years'] <- 10
age[suicide$age == '15-24 years'] <- 20
age[suicide$age == '25-34 years'] <- 30
age[suicide$age == '35-54 years'] <- 45
age[suicide$age == '55-74 years'] <- 65
age[suicide$age == '75+ years'] <- 85

suicide <- suicide %>%
  select(-one_of('age', 'sex')) %>%
  add_column(age, sex) %>%
  rename(gdp_year = gdp_for_year...., 
         gdp_per_capita = gdp_per_capita...., 
         country = ï..country, 
         hdi_year = HDI.for.year)

suicide$suicide_year <- buffer
suicide$suicide_year_female <- buffer
suicide$suicide_year_male <- buffer
suicide$population_year <- buffer

for (country_year in unique(suicide$country.year)) {
  data_curr <- suicide[suicide$country.year == country_year,]
  
  suicide_year_curr <- sum(data_curr$suicides_no)
  suicide_year_female_curr <- sum(data_curr[data_curr$sex == 0,]$suicides_no)
  suicide_year_male_curr <- sum(data_curr[data_curr$sex == 1,]$suicides_no)
  population_year_curr <- sum(data_curr[data_curr$sex == 1,]$population)
  
  suicide[suicide$country.year == country_year,]$suicide_year <- suicide_year_curr
  suicide[suicide$country.year == country_year,]$suicide_year_female <- suicide_year_female_curr
  suicide[suicide$country.year == country_year,]$suicide_year_male <- suicide_year_male_curr
  suicide[suicide$country.year == country_year,]$population_year <- population_year_curr
}

suicide$gdp_year = as.numeric(gsub(",", "", suicide$gdp_year))
suicide$sex = as.factor(suicide$sex)
suicide$age = as.factor(suicide$age)
suicide$generation = as.factor(suicide$generation)

head(suicide)
```

## Pre-analysis

In the last part, we found that HDI could be replaced by log(GDP), since the linear model with the response of HDI and predictor of log(GDP) has a R-squared value over 0.8. Considering the HDI variables contains quite some number of NA, we decided to use log(GDP) in the later analysis.
```{r}
hist_suicide_year <- ggplot(suicide) +
  geom_histogram(aes(x = suicide_year, y = ..density..), binwidth = 100) +
  theme_light()

hist_suicide_100k <- ggplot(suicide) +
  geom_histogram(aes(x = suicides.100k.pop, y = ..density..), binwidth = 1) +
  theme_light()

pie_sex <- ggplot(suicide, aes(x = "", y = suicides_no, fill = sex)) +
  geom_bar(stat="identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()

pie_age <- ggplot(suicide, aes(x = "", y = suicides_no, fill = age)) +
  geom_bar(stat="identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()

pie_gen <- ggplot(suicide, aes(x = "", y = suicides_no, fill = generation)) +
  geom_bar(stat="identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void()

df_suicide_gdp <- suicide %>%
  distinct(country.year, .keep_all = TRUE)

point_gdp_cap <- ggplot(df_suicide_gdp) +
  geom_point(aes(x = gdp_per_capita, y = suicide_year))

point_gdp_year <- ggplot(df_suicide_gdp) +
  geom_point(aes(x = gdp_year, y = suicide_year))

list_year = sort(unique(df_suicide_gdp$year))
suicide_by_year <- c()
for (year in list_year) {
  suicide_by_year[length(suicide_by_year) + 1] <- sum(df_suicide_gdp[df_suicide_gdp$year == year,]$suicide_year)
}

f_trend <- ggplot() +
  geom_line(aes(x = list_year, y = suicide_by_year))

ggarrange(hist_suicide_year, hist_suicide_100k, ncol = 2)
ggarrange(pie_sex, pie_age, pie_gen, ncol = 3)
ggarrange(point_gdp_cap, point_gdp_year, f_trend, ncol = 3)

suicide %>%
  select(one_of('year', 'suicide_no', 'population', 'suicides.100k.pop', 'gdp_year', 'gdp_per_capita')) %>%
  pairs(col = 'darkorange')

plot(log(suicide$gdp_per_capita), suicide$hdi_year)
m_hdi_gdp <- lm(hdi_year ~ log(gdp_per_capita), data = suicide)
summary(m_hdi_gdp)$adj.r.squared
```


## Modelling
```{r}
m0_suicide <- lm(suicides.100k.pop ~ population + gdp_year + age + sex + log(gdp_per_capita), data = suicide)

m0_suicide_aic <- step(m0_suicide, trace = 0)

m1_suicide <- lm(suicides.100k.pop ~ (population + gdp_year + age + sex + log(gdp_per_capita)) ^ 5, data = suicide)

m1_suicide_aic <- step(m1_suicide, trace = 0)

m2_suicide <- lm(suicides.100k.pop ~ (population + gdp_year + age + sex + log(population) + log(gdp_year) + log(age)) ^ 2, data = suicide)

m2_suicide_aic <- step(m2_suicide, trace = 0)

m3_suicide <- lm(suicides.100k.pop ~ I(population ^ -1) + gdp_year + age + sex + log(gdp_per_capita), data = suicide)

m3_suicide_aic <- step(m0_suicide, trace = 0, k = log(n), direction = 'backward')

model1 = lm(suicides.100k.pop ~ hdi_year + gdp_year + generation + age + sex + population, data = suicide)
selected1 = step(model1, direction = "backward", trace = FALSE)

#so far it is the best one
model2 = lm(suicides.100k.pop ~ hdi_year * gdp_year * generation * age * sex * population, data = suicide)
selected2 = step(model2, direction = "backward", trace = FALSE)

model3 = lm(suicides.100k.pop ~ gdp_per_capita * age * sex, data = suicide )
selected3 = step(model3, direction = "backward", trace = FALSE)

model4 = lm(suicides.100k.pop ~ sex * age * gdp_per_capita * generation, data = suicide)
selected4 = step(model4, direction = "backward", trace = FALSE)

results_all <- rbind(get_all(m0_suicide),
                     get_all(m0_suicide_aic),
                     get_all(m1_suicide),
                     get_all(m1_suicide_aic),
                     get_all(m2_suicide),
                     get_all(m2_suicide_aic),
                     get_all(m3_suicide),
                     get_all(m3_suicide_aic),
                     get_all(model2))

results_all
vif(m0_suicide)
```



```{r}


qq_plot(m3_suicide)
```


# Results

## Tables and plots of the models

# Discussion

# Appendix





